---
title: "Bioinformatics Class 8"
author: "Barry Grant"
date: "4/26/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# K-means ckustering

Lets make up some data to try K-means on:

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```

Our tasks
 - Use the kmeans() function setting k to 2 and nstart=20 

```{r}
km <- kmeans(x, centers = 2, nstart=20)
km
```

```{r}
km$cluster
```



Inspect/print the results 

Q. How many points are in each cluster?

```{r}
km$size
```

Q. What ‘component’ of your result object details 
      - cluster size?
      - cluster assignment/membership? 
      - cluster center?

```{r}
# cluster assignment/membership? 
km$cluster
```

```{r}
# cluster center
km$centers
```



Plot x colored by the kmeans cluster assignment and 
      add cluster centers as blue points

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15)
```
      
      
      
 Q. Repeat for k=3, which has the lower tot.withinss? 
```{r}
km3 <- kmeans(x, centers = 3, nstart=20)
km3
```

```{r}
km3$tot.withinss
```


# Hierarchical clustering

Lets try on our x points

```{r}
# First we need to calculate point (dis)similarity
#   as the Euclidean distance between observations 
dist_matrix <- dist(x) 
```

Lets look at this distance matrix

```{r}
class(dist_matrix)
```


```{r}
dim(as.matrix(dist_matrix))

```



```{r}
# The hclust() function returns a hierarchical 
#  clustering model
hc <- hclust(d = dist_matrix)

# the print method is not so useful here 
hc  
```

Lets draw the tree

```{r}
plot(hc)
```

lets draw the tree 

# Draws a dendrogram

```{r}
plot(hc)
abline(h=6, col="red")
cutree(hc, h=6) # Cut by height h
```

I can cut the tree at any height to give our clusters..

```{r}
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
plot(x)
```

```{r}
col <- as.factor( rep(c("c1","c2","c3"), each=50) ) 
```

```{r}
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
col <- as.factor( rep(c("c1","c2","c3"), each=50) ) 
```
```{r}
plot(x, col=col)
```


## Q. Use the dist(), hclust(), plot() and cutree()
# functions to return 2 and 3 clusters
# Q. How does this compare to your known 'col' groups?
```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)

```

```{r}
#to return 2 clusters i could cut at h=2.5 or set k=2
members2 <- cutree(hc, k=2)

#
members3 <- cutree(hc, k=3)

plot(x, col=members3, pch=15)
```
#PCA 
#lets make some data 
```{r}
## Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)
dim(mydata)

## Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="") 
View(mydata)
```

```{r}
## Lets label the first 5 columns wt1, wt2, wt3, wt4 and wt5
## and the last 5 ko1, ko2 etc. to ko5 (for "knock-out")
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
 paste("ko", 1:5, sep="") ) 
View(mydata)
```


```{r}
## Fill in some fake read counts
for(i in 1:nrow(mydata)) {
 wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
 ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))

 mydata[i,] <- c(wt.values, ko.values)
}
head(mydata)
```

```{r}
#head(t(mydata))
```

```{r}
pca <- prcomp( t(mydata), scale = TRUE)
pca
```
# Lets plot our PCA 
```{r}
plot(pca$x[,1], pca$x[,2])
```

```{r}
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

Plot our scree plot
```{r}
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```

```{r}
#colour up our PCA plot
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
```

```{r}
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
identify(pca$x[,1], pca$x[,2], labels=colnames(mydata))
```

```{r}
## Click to identify which sample is which
identify(pca$x[,1], pca$x[,2], labels=colnames(mydata)) 


```
```{r}
#Check
colnames(mydata)
pca$x[,2]
```










